Incorporating periodic variability in hidden Markov models for animal movement

### 2016-07-12 paper3.Rnw
working on paper, leftover isec talk changes

\doublespacing
<<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
library("ggplot2"); theme_set(theme_bw())
scale_colour_discrete <- function(...,palette="Set1")
    scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Set1")
    scale_fill_brewer(...,palette=palette)
zmargin <- theme(panel.margin=grid::unit(0,"lines"))
library("knitr")
opts_chunk$set(fig.width=6,fig.height=4)
library("depmixS4")
library("plyr")
library("dplyr")
library("reshape2")
library("gridExtra")
library("RColorBrewer")
# if (!require("ggalt")) {
#     cat("installing ggalt from BB repo ...\n")
#     library("devtools")
#     install_github("bbolker/ggalt")
#     library("ggalt")
# }
# library("ggalt")
@

@@

Clustering time-series data into discrete groups can improve
prediction and provide insight into the nature of underlying,
unobservable states of the system. However, temporal variation in the
rates at which individuals move between groups can obscure
such signals. We use finite mixture and hidden Markov models (HMMs), two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers (\emph{Puma concolor coryi}).
Allowing for temporal
heterogeneity in transition probabilities, a straightforward but
rarely extension of the standard HMM framework, 
resolves some shortcomings of current
models and clarifies the behavioural patterns of
panthers. More generally, we point out that model misspecification
(omitting important sources of variation) can lead to overfitting,
and as a corollary that incorporating previously neglected
structure in statistical models can allow more
accurate assessment of appropriate model complexity.

@@

Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines\todo{citation?}, could be useful directions for future exploration.

@@

Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity. The number of free parameters in an HMM can be generalized by summing up the number of free parameters of the three distinct components. Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the parameters of the prior distribution, transition function and emission distributions. Then the number of free parameters of an HMM is:

\begin{equation}
\underbrace{k_{i}*(n-1)}_{\text{Initial}} 
{+} \underbrace{k_{t}*n*(n-1)}_{\text{Transition}}
{+} \underbrace{ k_{e}*n}_{\text{Emission}}
\label{eq:numparam}
\end{equation}

When increasing in number of states, the number of free parameters in time-homogeneous FMMs and HMMs and FMMs with temporal heterogenity will increase linearly, whereas, HMMs with temporal heterogenity will not by Eq.~\ref{eq:numparam}. When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.

We used simulations to predict hourly step length and ACF 
because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex. 
We used this approach to validate our models and comparing these models with the observed movements instead of the standard Viterbi predictions by the Viterbi algorithm because Viterbi predictions \emph{double count} the observed data. The Viterbi algorithm solves/predicts for the most probable sequence of movement states based on the observations \citep{zucchini_hidden_2009,langrock_flexible_2012}. It is useful to predict missing data in the observation sequence, but because it is conditional on the observed values, it can not relibly evaluate goodness of fit for the different structral complexities of HMM models.

@@

We simulated a two-state HMM with sinusoidal temporal transitions 100 times and fitted it with two to five state HMMs and without temporal transition.  Figure~\ref{fig:sim_results} shows heterogeneous transition models can always predict the correct number of states, whereas, can over estiate the number of states via BIC-optimal approach.

@@

<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters (\\textbf{todo: change scales from log() to log-scale; need a real/more complete figure caption here; emphasize that dashes are for emphasis only; rename models (time-homog/heterog)}">>=
# g0 <- ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
#     geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)+zmargin
# (g0c <- g0 +
#     geom_encircle(data=subset(all_cat_pars,
#                               (catID=="15" & log_mean<1.5) |
#                               (catID!="15" & log_mean<1.8)),
#                   colour="black",linetype=2,expand=0.05))
@

@@

The hardest problem with multiple latent states is interpreting them
biologically. We have no way of knowing what panthers are actually
thinking (it is certainly more complex than being in one of a small
number of discrete latent states); 
we don't know the ``true'' number of latent states,
nor are we able to observe them directly, although
incorporating additional direct observations of 
behaviour (if available) can at least partially
address this problem \citep{fryxell_multiple_2008}.
Three distinct
movement states seem biologically interpretable for Florida panthers
according to \citet{kerk2015hidden}: Short step length suggests resting
states, intermediate step length a foraging state, and long
step length a traveling state. 
\todo{do we need this?}
Figure~\ref{state_plots} compares the estimated parameter values
for several cats
(mean and standard deviation of the step length in each state) between
the time-homogeneous and time-heterogeneous models. 
In general, the states with longer mean step lengths are relatively
similar between model classes. For cats 14 and 15, the states with
the longest or next-longest mean step lengths have similar means
and standard deviations; for cats 1 and 2, three long-step states
in the homogeneous HMM appear to divide two long-step states in 
the heterogeneous HMM. For short-step states, the heterogeneous HMM
tends to identify a high-variance state, while the homogeneous HMM
picks up states with very short step lengths (questionable in any
case because we have not taken any special efforts to account for
GPS error).

@@

Identifying behavioural states based on observations is a
common methodological problem in behavioural ecology; finding
solutions that are both biologically and statistically interpretable
and defensible is challenging. Models that suggest a higher number of
states fit the data well but are hard to interpret biologically,
while 
models that are biologically interpretable are not always statistically
supported. While the HMM framework has been applied in many research
areas, very few studies have considered modeling with more than 3
states. HMM is a simple, straightforward framework, but HMM extensions
of adding a small amount of complexity can complicate the model
greatly and reveal deeper understanding. For example,
\citet{kerk2015hidden} used 2 to 6 movement states HSMM to model
Florida panthers, \citet{morales_extracting_2004} used habitat
distance dependent transition HMM to model elk, and outside of
ecology, \citet{raffa_multivariate_2015} used mixed effect transition
and emission HMM in a smoking study.

We have presented a relatively simple concept yet little-studied
extension by covariate dependent HMM can overfit the number of true
states in covariate dependent HMM (in our case, time-dependent
transitions). HMM with more than three states are not biologically
interpretable and often discarded, but we cannot ignore that
increasing complexity (number of states) states can increase goodness
of fit statistically. Our analyses revealed modeling temporal
heterogeneity in HMM transition can reduce the number of states
estimated in time homogeneous transitions HMM movement of Florida
panthers and reduce the overall model complexity.

For some kinds of model predictions, the time homogeneous transition
model can perform as good as the time inhomogeneous transition models
in terms of predicting the states by the Viterbi algorithm (the
tradition state prediction method) and can easily overlook more
complicated models. This approach allows the time homogeneous transition model's prediction to capture the temporal heterogeneity from the observations. The limitation of the Viterbi appoach is that it cannot predict future movements beyond the observation sequence. In general, the time homogeneous
transition model gives the broader scope of movement pattern where as
the time inhomogeneous transitions models gives a more realistic
movement structure and future predictions.

Although the time inhomogeneous transitions offer an objective
approach to (1) reduce the number of states to a biologically
interpretable level, (2) captures the observational autocorrelation
and temporal pattern, (3) reduce the overall complexity as illustrated
by our study, several challenges must be overcome in order to maximize
its potential. The current model is based on the reduced univariate
step length response emission without fitting turning angles. Although
turning angles are strongly correlated with step length, it does not
effect the overall prediction on number of states. But movement
trajectories predictions, and spatial models still require both step
length and turning angles. Furthermore, modelling temporal
heterogeneity only reduce the number of predicted states by 1
resulting in 4/5 states might suggest (1) other factors such as
spatial components that might also contribute to over predicting
number of states, or (2) there really are 1 or 2 more different
movement characteristics that are biologically unknown. Despite the
challenges, our study provides an important extension in HMM modeling
that is not in the seen before and a real life application that it
exist and applicable in animal movement ecology.

@@

\todo{does it make any sense to compress figs~\ref{adj_BIC-comparisons_cat2} etc. into a single plot (via facetgrid or multi-lines)?  Same idea for all repeated plots; also, update to use percent-+-percent syntax}

>>>>

\doublespacing
<<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
library("ggplot2"); theme_set(theme_bw())
scale_colour_discrete <- function(...,palette="Set1")
    scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Set1")
    scale_fill_brewer(...,palette=palette)
zmargin <- theme(panel.margin=grid::unit(0,"lines"))
library("knitr")
opts_chunk$set(fig.width=6,fig.height=4)
## fix complaints about "incompatible color definition" from xcolor ...
## http://tex.stackexchange.com/questions/148188/knitr-xcolor-incompatible-color-definition
knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)})
library("depmixS4")
library("plyr")
library("dplyr")
library("reshape2")
library("gridExtra")
library("RColorBrewer")
# if (!require("ggalt")) {
#     cat("installing ggalt from BB repo ...\n")
#     library("devtools")
#     install_github("bbolker/ggalt")
#     library("ggalt")
# }
library("ggalt")
@

@@

Clustering time-series data into discrete groups can improve
prediction and provide insight into the nature of underlying,
unobservable states of the system. However, temporal variation in the
rates at which individuals move between groups can obscure
such signals. We use finite mixture and hidden Markov models (HMMs), two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers (\emph{Puma concolor coryi}).
Allowing for temporal
heterogeneity in transition probabilities, a straightforward but
little-used extension of the standard HMM framework, 
resolves some shortcomings of current
models and clarifies the behavioural patterns of
panthers. More generally, we point out that model misspecification
(omitting important sources of variation) can lead to overfitting,
and as a corollary that incorporating previously neglected
structure in statistical models can allow more
accurate assessment of appropriate model complexity.

@@

Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines, could be useful directions for future exploration.

@@

Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity. The number of free parameters in an HMM can be generalized by summing up the number of free parameters of the three distinct components. Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$. Then the number of free parameters of an HMM is:

\begin{equation}
\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
{+} \underbrace{k_{t}\cdot n\cdot (n-1)}_{\text{Transition}}
{+} \underbrace{ k_{e}\cdot n}_{\text{Emission}}
\label{eq:numparam}
\end{equation}

As the number of states increases, the number of free parameters in time-homogeneous FMMs and HMMs and FMMs with temporal heterogenity will increase linearly, whereas HMMs with temporal heterogenity will increase quadratically (Eq.~\ref{eq:numparam}). When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.

We used simulations to predict hourly step length and ACF 
because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex. 
We used this approach to validate our models and comparing these models with the observed movements instead of the standard Viterbi predictions by the Viterbi algorithm because Viterbi predictions, which use the most probable sequence of movement states based on the observations \citep{zucchini_hidden_2009,langrock_flexible_2012}, double-count the observed data. It is useful to predict missing data in the observation sequence, but because it is conditional on the observed values, it can not reliably evaluate goodness of fit for the different structural complexities of HMM models. \todo{think about/check out \textbf{pseudo-residuals}}

@@

\todo{The description of the simulations should be in Methods, not Results; here we should describe the results of the simulations only}
We simulated a two-state HMM with sinusoidal temporal transitions 100 times and fitted it with two to five state HMMs and without temporal transition.  Figure~\ref{fig:sim_results} shows heterogeneous transition models can always predict the correct number of states, whereas, can overestimate the number of states via BIC-optimal approach.

@@

<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="Step-length parameters (\\textbf{todo: change scales from log() to log-scale; need a real/more complete figure caption here; emphasize that dashes are for emphasis only; rename models (time-homog/heterog)}">>=
g0 <- ggplot(all_cat_pars,aes(log_mean,log_sd,color=Model))+
     geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)+zmargin
 (g0c <- g0 +
     geom_encircle(data=subset(all_cat_pars,
                               (catID=="15" & log_mean<1.5) |
                               (catID!="15" & log_mean<1.8)),
                   colour="black",linetype=2,expand=0.05))
@

@@

The hardest problem with multiple latent states is interpreting them
biologically. We have no way of knowing what panthers are actually
thinking (it is certainly more complex than being in one of a small
number of discrete latent states); 
we don't know the ``true'' number of latent states,
nor are we able to observe them directly, although
incorporating additional direct observations of 
behaviour (if available) can at least partially
address this problem \citep{fryxell_multiple_2008}.
Three distinct
movement states seem biologically interpretable for Florida panthers
according to \citet{kerk2015hidden}: Short step length suggests resting
states, intermediate step length a foraging state, and long
step length a traveling state. 
\todo{do we need this?}
Figure~\ref{fig:state_plots} compares the estimated parameter values
for several cats
(mean and standard deviation of the step length in each state) between
the time-homogeneous and time-heterogeneous models. 
In general, the states with longer mean step lengths are relatively
similar between model classes. For cats 14 and 15, the states with
the longest or next-longest mean step lengths have similar means
and standard deviations; for cats 1 and 2, three long-step states
in the homogeneous HMM appear to divide two long-step states in 
the heterogeneous HMM. For short-step states, the heterogeneous HMM
tends to identify a high-variance state, while the homogeneous HMM
picks up states with very short step lengths (questionable in any
case because we have not taken any special efforts to account for
GPS error).

@@

Identifying behavioural states based solely on
observations of movement behaviour is a
common methodological problem in behavioural ecology; finding
solutions that are both biologically interpretable
and statistically defensible is challenging. 
Models with large numbers of
states fit the data well (even when model complexity
is penalized) but are hard to interpret biologically,
while models with fewer states
that are biologically interpretable are not always statistically
supported. While the HMM framework has been applied in many research
areas, few studies have considered modeling with more than 3
states. \todo{what is the rest of this paragraph supposed to get across?}
HMM is a simple, straightforward framework, but HMM extensions
of adding a small amount of complexity can complicate the model
greatly and reveal deeper understanding. For example,
\citet{kerk2015hidden} used 2 to 6 movement states HSMM to model
Florida panthers, \citet{morales_extracting_2004} used habitat
distance dependent transition HMM to model elk, and outside of
ecology, \citet{raffa_multivariate_2015} used mixed effect transition
and emission HMM in a smoking study.

We have presented a relatively simple yet little-studied
extension by covariate dependent HMM can overfit the number of true
states in covariate dependent HMM (in our case, time-dependent
transitions). HMM with more than three states are not biologically
interpretable and often discarded, but we cannot ignore that
increasing complexity (number of states) states can increase goodness
of fit statistically. Our analyses revealed modeling temporal
heterogeneity in HMM transition can reduce the number of states
estimated in time homogeneous transitions HMM movement of Florida
panthers and reduce the overall model complexity.

For some kinds of model predictions, the time homogeneous transition
model can perform as well as the time inhomogeneous transition models
in terms of predicting the states by the Viterbi algorithm (the
tradition state prediction method) and can easily overlook more
complicated models. This approach allows the time homogeneous
transition model's prediction to capture the temporal heterogeneity
from the observations. The limitation of the Viterbi appoach is that
it cannot predict future movements beyond the observation sequence. In
general, the time homogeneous transition model gives the broader scope
of movement pattern where as the time inhomogeneous transitions models
gives a more realistic movement structure and future predictions.

Models with time-inhomogeneous transitions appear to
offer a simple way to (1) reduce the selected number of states to a biologically
interpretable level, (2) capture observed diurnal and autocorrelation
patterns, (3) provide fitted models that are better (lower BIC)
and, overall, less complex than time-homogeneous models.
However, several challenges must be overcome in order to maximize
their potential. The current model is based on the reduced univariate
step length response emission without fitting turning angles. Although
turning angles are strongly correlated with step length, it does not
affect the overall prediction on number of states. But movement
trajectories predictions, and spatial models still require both step
length and turning angles. Furthermore, modelling temporal
heterogeneity only reduce the number of predicted states by 1
resulting in 4/5 states might suggest (1) other factors such as
spatial components that might also contribute to over predicting
number of states, or (2) there really are 1 or 2 more different
movement characteristics that are biologically unknown. Despite the
challenges, our study provides an important extension in HMM modeling
that has not been seen before and a real life-application to
animal movement ecology.

@@

\todo{does it make any sense to compress figs~\ref{fig:adj_BIC_comparisons_cat2} etc. into a single plot (via facetgrid or multi-lines)?  Same idea for all repeated plots; also, update to use percent-+-percent syntax}
